---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading packages
First, let's load all packages that we make use of for this assignment.
```{r, load_packages, message=FALSE}
library(tidyverse)
```
FOR OUTPUT MANIPULATION
## Loading and preprocessing the data
UNZIPPED DATA ARE READ FROM HERE
```{r, read_data}
data <- read_csv("activity.csv")
str(data)
```
The `read_csv` command automatically recognizes the variables as containing double and data values. There does not seem to be a need to further tidy up the data.

## What is mean total number of steps taken per day?
As the data is provided for 5-minute intervalls, we first have to sum up the steps taken for each day before we can plot the data.
```{r, daily_steps1, message=FALSE, warning=FALSE}
daily <- data %>% group_by(date) %>% 
    summarise(StepsPerDay = sum(steps))

ggplot(data = daily, aes(x = StepsPerDay)) + 
    geom_histogram() + 
    labs(title = "Histogramm of Average Steps per Day", x = "Steps Per Day")
```

The histogramm shows that, on most days, around 10000 steps have been recorded.

This is underlined when looking at summary statistics.

```{r, daily_steps2}
summary(daily$StepsPerDay)
```
A PLAIN SUMMARY OF STEPS PER DAY

## What is the average daily activity pattern?
In order to find a daily pattern we can look at the average steps taken, grouped by the provided 5-minute intervalls.

```{r, daily_pattern, message=FALSE, warning=FALSE}
pattern <- data %>% group_by(interval) %>% 
    summarise(StepsPerInt = mean(steps, na.rm = TRUE))

ggplot(data = pattern, aes(x = interval, y = StepsPerInt)) + 
    geom_line() +
    labs(title = "Average Steps Taken Over The Course of a Day", 
        x = "5-Minute Interval", y = "Average Number of Steps per Interval")
```

A SPIKE IS OBSERVABLE AT 800

```{r, daily_pattern2, message=FALSE, warning=FALSE}
max(pattern$StepsPerInt)
```
The maximum number of average steps taken for a 5-minute intervall is 206.

```{r, daily_pattern3, message=FALSE, warning=FALSE}
pattern$interval[which(pattern$StepsPerInt == max(pattern$StepsPerInt))]
```
The interval associated with this value is interval number 835.


## Imputing missing values
```{r, missings}
sum(is.na(data$steps))
```

FILLING THE 2304 MISSING VALUES IN THE DATASET

```{r, mean_function}
getIntervalMean <- function(interval) {
    pattern[pattern$interval == interval,]$StepsPerInt
}
```

Now we can simply loop over all rows and replace missing values with the mean number of steps for the respective interval. We will store the complete dataset in a new object.

```{r, fill_na}
data_complete <- data
for (i in 1:length(data_complete$steps)) {
    if(is.na(data_complete$steps[i])) {
        data_complete$steps[i] <- getIntervalMean(data_complete$interval[i])
    }
}
```
Let's inspect the filled in dataset...


```{r, daily_steps_imputed, message=FALSE, warning=FALSE}
daily_imputed <- data_complete %>% group_by(date) %>% 
    summarise(StepsPerDay = sum(steps))

ggplot(data = daily_imputed, aes(x = StepsPerDay)) + 
    geom_histogram() + 
    labs(title = "Histogramm of Average Steps per Day", x = "Steps Per Day")
```
```{r, daily_steps_imputed2}
summary(daily_imputed$StepsPerDay)
```
THE PLOT AND SUMMARY HAS ALMOST SAME MESSAGE

## Are there differences in activity patterns between weekdays and weekends?

ADDING A FACTOR VARIABLE TO DETECT A PATTERN

```{r, weekday}
weekend_fun <- function(weekday) {
    if(weekday %in% c("mondays", "tuesdays", "wednesdays", "thursdays", "fridays")) {
        return(0)
    }
    else if(weekday %in% c("egwuregwu", "ubosi uka")) {
        return(1)
    }
}

data_complete$weekend <- sapply(weekdays(data_complete$date), weekend_fun)
data_complete$weekend <- factor(data_complete$weekend)
```

we compare the weekday and weekend patterns

```{r, patterns_weekend, message=FALSE, warning=FALSE}
pattern_weekend <- data_complete %>% group_by(weekend,interval) %>% 
    summarise(StepsPerInt = mean(steps, na.rm = TRUE))

ggplot(data = pattern_weekend, aes(x = interval, y = StepsPerInt)) + 
    geom_line() +
    facet_grid(rows = vars(weekend)) +
    labs(title = "Average Steps Taken Over The Course of a Day", 
        x = "5-Minute Interval", y = "Average Number of Steps per Interval")
```

FROM THE PATTERN ANALYSIS, STEPS ARE MORE DURING WEEKDAYS THAN WEEKENDS
UKACHUKWU CHRISTIAN_SPIRITUAL